---
---
---

# Using DEP for Volcano Plots

```{r}
library("DEP")
library("dplyr")
library(readr)
library(tidyverse)
library(glue)
```

## Your Files

Make sure both your `combined_proteins` and `experiment_annotations` files are placed in the `data` folder. Once that's done, you can just paste in the **file names** into the two variables below:

```{r}
combined_proteins_file="INSERT_YOUR_FILE_PATH_HERE"
experiment_annotation_file="INSERT_YOUR_FILE_PATH_HERE"


## Example used for this notebook:
combined_proteins_file="combined_protein_reMDAMB_fixed.tsv"
experiment_annotation_file="experiment_annotation_reMDAMB_fixed.tsv"

```

## Reading data in

```{r}
proteins_raw <- read_tsv(file = paste0("../data/", combined_proteins_file)) |> as.data.frame()

metadata_raw <- read_tsv(file = paste0("../data/", experiment_annotation_file)) |> as.data.frame()

```

## Data Formatting

```{r}
# Remove Contamination
proteins_raw_remove_contam <- proteins_raw |> filter(!grepl("contam", Protein))

# Select the important columns
proteins_shorten_raw <- select(proteins_raw_remove_contam, Protein, "Protein ID", "Entry Name", Description, contains("LFQ"))

# Remove spaces from column names
# "make.names" is a special function that formats all column names
proteins <- proteins_shorten_raw |> rename_with(make.names)
```

## Extract protein name from `Entry.Name`

```{r}
## Split the Protein name by delimiter "|"
proteins <- separate_wider_delim(proteins, cols=Protein, delim = "|", names = c("first", "second", "third"))

# Split again to remove the "HUMAN" part of "XXX_HUMAN"
proteins <- separate_wider_delim(proteins, cols = third, delim = "_", names= c("name", "human"))

# Remove other columns that we made during the process
proteins <- proteins %>% select(-c("first", "second", "human"))
```

```{r}
# Use DEP's function to prepare final version
proteins_for_dep <- make_unique(proteins, 
                                  names="name", 
                                  ids="Protein.ID")
```

## Make SummarizedExperiment Object

### Prepare LFQ Column Numbers for DEP

```{r}
# DEP needs the column numbers that actually have the LFQ intensities
LFQ_columns <- grep("LFQ", colnames(proteins_for_dep))
```

### Prepare Metadata for DEP

```{r}
# Remove columns we don't need 
metadata_for_dep = metadata_raw |> select(-c(file, sample))

# Rename columns, since DEP is expecting only three columns:
#  "label", "condition", "replicate"
metadata_for_dep = metadata_for_dep |> rename("label" = "sample_name")
```

### Make the SummarizedExperiment Object

```{r}
# Use DEP to make a SummarizedExperiment (se) object
data <- make_se(proteins_for_dep, LFQ_columns, metadata_for_dep)
```

## Visualizations of Data

```{r}
plot_frequency(data)
```

```{r}
plot_numbers(data)
```

```{r}
plot_coverage(data)
```

```{r}
data_normalized <- normalize_vsn(data)
```

```{r}
plot_normalization(data, data_normalized)
```

```{r}
plot_missval(data)
```

```{r}
data_imputed <- impute(data_normalized, fun = "MinProb", q = 0.01)
plot_imputation(data_normalized, data_imputed)
```

```{r}

```

## Differential Enrichment Analysis

Contrasts need to be defined in the format of `CONDITION1_vs_CONDITION2` , and they need to match the name in the `condition` column in `metadata_for_dep`. For example, if you want to compare the condition `"control123"` with `"wildtype456"`, then the manual contrast will be `"control123_vs_wildtype456"`.

```{r}
# Manually define contrasts for the volcano plot
data_contrasts <- test_diff(data_imputed, type = "manual", 
                              test = c("control468_vs_low468",
                                       "control231_vs_high231"))
```

## Make DEP object

`alpha`: The threshold for the adjusted p-value. Here, a sample value of `0.05` is used.

`lfc`: the threshold for the log2 fold change. Here, a sample value of `1.5` fold change (then logged)

```{r}
dep <- add_rejections(data_contrasts, alpha = 0.05, lfc = log2(1.5))
```

## Visualizations for DEP objects

```{r}
plot_pca(dep, x = 1, y = 2, n = 500, point_size = 4)
```

```{r}
plot_cor(dep, significant = TRUE, lower = 0, upper = 1, pal = "Spectral", font_size = 4)
```

```{r}
plot_heatmap(dep, type = "centered", kmeans = TRUE,  col_limit = 4,
             show_row_names = FALSE, indicate = c("condition", "replicate"),
             col_font_size = 6)
```

```{r}
plot_heatmap(dep, type = "contrast", kmeans = TRUE, col_font_size = 5,
             k = 6, col_limit = 10, show_row_names = FALSE)
```

### Volcano Plots for Contrasts

```{r}
plot_volcano(dep, contrast = "control468_vs_low468", label_size = 2, add_names = TRUE)
```

```{r}
plot_volcano(dep, contrast = "control231_vs_high231", label_size = 2, add_names = TRUE)
```
