---
---
---

# Using DEP for Volcano Plots

```{r}
library("DEP")
library("dplyr")
library(readr)
library(tidyverse)
library(glue)
library(pcaMethods)
```

## Your Files

Make sure both your `combined_proteins` and `experiment_annotations` files are placed in the `data` folder. Once that's done, you can just paste in the **file names** into the two variables below:

```{r}
combined_proteins_file="INSERT_YOUR_FILE_PATH_HERE"
experiment_annotation_file="INSERT_YOUR_FILE_PATH_HERE"


## Example used for this notebook:
combined_proteins_file="combined_protein_reMDAMB_fixed.tsv"
experiment_annotation_file="experiment_annotation_reMDAMB_fixed.tsv"

```

## Reading data in

```{r}
proteins_raw <- read_tsv(file = paste0("../data/", combined_proteins_file)) |> as.data.frame()

metadata_raw <- read_tsv(file = paste0("../data/", experiment_annotation_file)) |> as.data.frame()

```

## Data Formatting

```{r}
# Remove Contamination
proteins_raw_remove_contam <- proteins_raw |> filter(!grepl("contam", Protein))

# Select the important columns
proteins_shorten_raw <- select(proteins_raw_remove_contam, Protein, "Protein ID", "Entry Name", Description, contains("LFQ"))

# Remove spaces from column names
# "make.names" is a special function that formats all column names
proteins <- proteins_shorten_raw |> rename_with(make.names)
```

## Extract protein name from `Entry.Name`

```{r}
## Split the Protein name by delimiter "|"
proteins <- separate_wider_delim(proteins, cols=Protein, delim = "|", names = c("first", "second", "third"))

# Split again to remove the "HUMAN" part of "XXX_HUMAN"
proteins <- separate_wider_delim(proteins, cols = third, delim = "_", names= c("name", "human"))

# Remove other columns that we made during the process
proteins <- proteins %>% select(-c("first", "second", "human"))
```

```{r}
# Use DEP's function to prepare final version
proteins_for_dep <- make_unique(proteins, 
                                  names="name", 
                                  ids="Protein.ID")
```

## Make SummarizedExperiment Object

### Prepare LFQ Column Numbers for DEP

```{r}
# DEP needs the column numbers that actually have the LFQ intensities
LFQ_columns <- grep("LFQ", colnames(proteins_for_dep))
```

### Prepare Metadata for DEP

```{r}
# Remove columns we don't need 
metadata_for_dep = metadata_raw |> select(-c(file, sample))

# Rename columns, since DEP is expecting only three columns:
#  "label", "condition", "replicate"
metadata_for_dep = metadata_for_dep |> rename("label" = "sample_name")
```

### Make the SummarizedExperiment Object

```{r}
# Use DEP to make a SummarizedExperiment (se) object
data <- make_se(proteins_for_dep, LFQ_columns, metadata_for_dep)
```

## Visualizations of Data

```{r}
plot_frequency(data)
```

```{r}
plot_numbers(data)
```

```{r}
plot_coverage(data)
```

```{r}
data_normalized <- normalize_vsn(data)
```

```{r}
plot_normalization(data, data_normalized)
```

```{r}
plot_missval(data)
```

```{r}
data_imputed <- impute(data_normalized, fun = "MinProb", q = 0.01)
plot_imputation(data_normalized, data_imputed)
```

```{r}

```

## Differential Enrichment Analysis

Contrasts need to be defined in the format of `CONDITION1_vs_CONDITION2` , and they need to match the name in the `condition` column in `metadata_for_dep`. For example, if you want to compare the condition `"control123"` with `"wildtype456"`, then the manual contrast will be `"control123_vs_wildtype456"`.

```{r}
# Manually define contrasts for the volcano plot
data_contrasts <- test_diff(data_imputed, type = "manual", 
                              test = c("low468_vs_control468",
                                       "high231_vs_control231",
                                       "drug231_vs_control231",
                                       "drug468_vs_control468",
                                       "high_drug231_vs_high231",
                                       "low_drug468_vs_low468",
                                       "low_drug468_vs_drug468",
                                       "high_drug231_vs_drug231"))
```


```{r}
design <- model.matrix(~ 0 + cell + gene + drug, data = metadata_raw)
fit <- lmFit(data_imputed, design)
con <- makeContrasts(drug_yes - drug_no, levels = design)
fit.con <- contrasts.fit(fit, con)
bayes <- eBayes(fit.con)
results <- topTable(bayes, n =Inf, adjust = "fdr")

```

## Make DEP object

`alpha`: The threshold for the adjusted p-value. Here, a sample value of `0.05` is used.

`lfc`: the threshold for the log2 fold change. Here, a sample value of `1.5` fold change (then logged)

```{r}
dep <- add_rejections(data_contrasts, alpha = 0.05, lfc = log2(1.5))
```

## Visualizations for DEP objects

```{r}
plot_pca(dep, x = 1, y = 2, n = 500, point_size = 4)
```

```{r}
plot_cor(dep, significant = TRUE, lower = 0, upper = 1, pal = "Spectral", font_size = 4)
```

```{r}
plot_heatmap(dep, type = "centered", kmeans = TRUE,  col_limit = 4,
             show_row_names = FALSE, indicate = c("condition", "replicate"),
             col_font_size = 6)
```

```{r}
plot_heatmap(dep, type = "contrast", kmeans = TRUE, col_font_size = 5,
             k = 6, col_limit = 10, show_row_names = FALSE)
```

### Volcano Plots for Contrasts

#### Does genetic engineer affect the overall proteome changes in 468 and 231?

```{r}
plot_volcano(dep, contrast = "low468_vs_control468", label_size = 2, add_names = TRUE)
```

```{r}
plot_volcano(dep, contrast = "high231_vs_control231", label_size = 2, add_names = TRUE)
```

```{r}
plot_volcano(dep, contrast = "drug231_vs_control231", label_size = 2, add_names = TRUE)
```

```{r}
plot_volcano(dep, contrast = "drug468_vs_control468", label_size = 2, add_names = TRUE)
```

```{r}
plot_volcano(dep, contrast = "high_drug231_vs_high231", label_size = 2, add_names = TRUE)
```

```{r}
plot_volcano(dep, contrast = "low_drug468_vs_low468", label_size = 2, add_names = TRUE)
```

#### Does the drug have a diffferent effect on control cell lines vs engineered cell lines?

```{r}
plot_volcano(dep, contrast = "low_drug468_vs_drug468", label_size = 2, add_names = TRUE)
```

```{r}
plot_volcano(dep, contrast = "high_drug231_vs_drug231", label_size = 2, add_names = TRUE)
```

```{r}
{r}
# NEED TO LOAD THIS LIBRARY FIRST
library("enrichR")
res <- listEnrichrDbs()

GO_result = test_gsea(dep, databases = c("GO_Biological_Process_2023"))

result = test_gsea(dep, databases = c("GO_Biological_Process_2023", "KEGG_2021_Human"))
```

```{r}
KEGG_result = test_gsea(dep, databases = "KEGG_2021_Human")
```

```{r}
reactome_result = test_gsea(dep, databases = "Reactome_2022")
```

```{r}
HDSigDB_result = test_gsea(dep, databases = "HDSigDB_Human_2021")
```

##low468_vs_control468

```{r}
low468_vs_control468_data = result |> filter(contrast == "low468_vs_control468")|> filter(P.value < 0.05)

plotEnrich(low468_vs_control468_data, showTerms = 30, numChar = 60)

```

```{r}
low468_vs_control468_data_kegg = KEGG_result |> filter(contrast == "low468_vs_control468")|> filter(P.value < 0.05)

plotEnrich(low468_vs_control468_data_kegg, showTerms = 30, numChar = 60)
```

```{r}
low468_vs_control468_data_HD = HDSigDB_result |> filter(contrast == "low468_vs_control468")

plotEnrich(low468_vs_control468_data_HD, showTerms = 30, numChar = 100)
```

##drug468 vs control468

```{r}
drug468_vs_control468_data = result |> filter(contrast == "drug468_vs_control468")|> filter(P.value < 0.05)

plotEnrich(drug468_vs_control468_data, showTerms = 30, numChar = 60)
```

```{r}
drug468_vs_control468_data_reac = reactome_result |> filter(contrast == "drug468_vs_control468")|> filter(P.value < 0.05)

plotEnrich(drug468_vs_control468_data_reac, showTerms = 30, numChar = 60)
```

```{r}
drug468_vs_control468_data_go = GO_result |> filter(contrast == "drug468_vs_control468")|> filter(P.value < 0.05)

plotEnrich(drug468_vs_control468_data_go, showTerms = 30, numChar = 60)

```

##drug231 vs control231

```{r}
drug231_vs_control231_data = result |> filter(contrast == "drug231_vs_control231") |> filter(P.value < 0.05)

plotEnrich(drug231_vs_control231_data, showTerms = 30, numChar = 60)
```

```{r}
library(enrichplot)
ridgeplot(drug231_vs_control231_data)
```

##high231 vs control231

```{r}
high231_vs_control231_data = result |> filter(contrast == "high231_vs_control231")|> filter(P.value < 0.05)

plotEnrich(high231_vs_control231_data, showTerms = 30, numChar = 60)

```

```{r}
high231_vs_control231_data_reac = reactome_result |> filter(contrast == "high231_vs_control231")|> filter(P.value < 0.05)

plotEnrich(high231_vs_control231_data_reac, showTerms = 70, numChar = 60)
```

##high_drug231 vs drug231

```{r}
drug_high231_vs_high231_data = result |> filter(contrast == "high_drug231_vs_high231")|> filter(P.value < 0.05)

plotEnrich(drug_high231_vs_high231_data, showTerms = 30, numChar = 60)
```

##drug_low468_vs_low468

```{r}
drug_low468_vs_low468_data = result |> filter(contrast == "low_drug468_vs_low468")|> filter(P.value < 0.05)

plotEnrich(drug_low468_vs_low468_data, showTerms = 30, numChar = 60)

```

```{r}
drug_low468_vs_drug468_data = result |> filter(contrast == "low_drug468_vs_drug468")|> filter(P.value < 0.05)

plotEnrich(drug_low468_vs_drug468_data, showTerms = 30, numChar = 60)
```

```{r}
drug_high231_vs_drug231_data = result |> filter(contrast == "high_drug231_vs_drug231")|> filter(P.value < 0.05)

plotEnrich(drug_high231_vs_drug231_data, showTerms = 30, numChar = 60)
```

```{r}

```

## Testing enrichment

```{r}
row_data = SummarizedExperiment::rowData(dep, use.names = FALSE)
```

```{r}
df <- row_data |> as.data.frame() |> select(name, ends_with("_significant")) |> mutate(name=gsub("[.].*", "", name))

contrast = "low_drug468_vs_low468_significant"

significant = df[df[[contrast]],]

genes = significant$name

enrich_result_test = enrichR::enrichr(genes, databases = "GO_Biological_Process_2023")
```

```{r}
organism = "org.Hs.eg.db"

```

```{r}
data(geneList, package="DOSE")
de <- names(geneList)[abs(geneList) > 2]
ego <- enrichGO(de, OrgDb = "org.Hs.eg.db", ont="BP", readable=TRUE)
```
